Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 introduction.tex
minor changes to gravity section

Many phenomena in astrophysics and cosmology require a high spatial and temporal dynamical range, and there are a number of numerical techniques that have been used to model such problems.  At the current time, the most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful -- particularly for those situations dominated by gravity, where systems evolve over large spatial and temporal dynamic ranges.  Its development, however, is still at a relatively early stage when compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.

Therefore, we would like to make use of the expertise that has been invested in Eulerian solvers; unfortunately, as originally developed, none of these schemes provide an easy way to adaptively and dynamically increase the spatial and temporal resolution in small volumes, as required by, for example, gravitational instability.  We have instead adopted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add and modify additional, finer meshes (``patches'') over regions that require higher resolution.  Since each patch has a simple geometry (Cartesian, in our case) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is relatively straightforward to add other advanced physics (e.g. magnetohydrodynamics, radiation transport, chemistry, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, MHD, radiation transport, star formation and a range of other physical effects.

There have been many other numerical methods described in the astronomical literature which contain elements of this idea, or have a similar aim.  For example, an N-body solver \citep{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions.  This static approach, applied to hydrodynamics, has been used extensively \citep[e.g.,][]{Ruffert94, Anninos94}.  Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes that possess this feature, both with and without  hydrodynamics  \citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}.  Of these, perhaps the most comparable and widely used are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells.   It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a few grid codes have adopted an unstructured approach based on a moving Voronoi mesh \citep{Arepo10, Tess11}.

\enzo\ is a structured adaptive mesh refinement (SAMR) code that was originally developed for cosmological hydrodynamics and has been used on a wide variety of problems.  It has been expanded as a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily expanded.  Although many of the components of the code have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N}, there has been no systematic and complete description of the code.  In this paper we provide that description, filling in many gaps and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, and more than 100 peer-reviewed papers have been published based on results obtained with the code.  The code has previously been used to study a wide range of astrophysical systems, including galaxy clusters \citep{Loken02}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01}, cooling flows \citep{Li12}, turbulence \citep{Kritsuk04}, and the formation of the first stars \citep{ABN02}.

Finally, we note that the numerical simulation of astronomical phenomena now plays a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01, DecadalSurvey10}.   But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

>>>>

Due to the high spatial and temporal dynamical ranges involved, astrophysics and cosmological phenomena present a taxing challenge for simulators. To tackle these, a number of numerical techniques have been developed which can be broadly split into the gridlesss,fully Lagrangian methods and grid-based, Eulerian schemes. The most commonly used is an example of the first typeand is known as Smoothed Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achived much success, particularly in regimes dominated by gravity.
%where systems evolve over large spatial and temporal dynamic ranges (cut because we've already said that in the first sentense - EJT)
However, its development to include an increasing number of sort-after physical processes is still at a relatively early stage when compared with the effort put into the latter type of Eulerian grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.
% is that what was meant by `` Its development, however, is still at a relatively early stage''? Since SPH is older than AMR, I found this a little confusing as written.

Despite this invested expertise, in their original form the Eulerian solvers have a serious drawback: they do not provide an easy method of adaptively increasing the spatial and temporal resolution in small volumes of the simulation. Such flexibility is essential for following physical processes such as gravitational instability. 
%Yet, despite the expertise invested in Eulerian solvers, in their original form they do not provide an easy method of adaptively increasing the spatial and temporal resolution in small volumes of the simulation. Such flexibility is essential for following physical processes such as gravitational instability. 
A solution to this was first proposed by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, and became known as Structured Adaptive Mesh Refinement (SAMR). The pricipal is to adaptively add and modify additional, finer meshes (``grids'') over regions that require higher resolution. In addition, it is possible to add other more advanced physics including --for the implementation in the astrophysics code presented in this paper-- comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, magnetohydrodynamics, radiation transport, star formation and a range of other physical effects.
% I was concerned the original text sounded a little like this was a brand new technique in a brand new code (despite the references below) which would be confusing to a reader familiar with the field and in particular, Enzo.

There have been a number of numerical methods described in the astronomical literature which contain elements of SAMR or have a similar aim. For example, the N-body solver developed by \citet{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions. This static approach was later used extensively when applied to hydrodynamics \citep[e.g.,][]{Ruffert94, Anninos94}. Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes that possess this feature, both with and without  hydrodynamics  \citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}. Of these, perhaps the most comparable and widely used are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells. It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a few grid codes have adopted an unstructured approach based on a moving Voronoi mesh \citep{Arepo10, Tess11}.

In this paper, we present \enzo\ ; a structured adaptive mesh refinement (SAMR) code. Originally developed for cosmological hydrodynamics, \enzo\ has since been used on a wide variety of problems.  It has grown to become a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily extended to include new capabilities.  Although many of the components of the \enzo\ have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N, WangAbelZhang08,ReynoldsHayesPaschosNorman2009,Collins10,Wise11_Moray}, there has previously been no systematic and complete description of the code.  In this paper we provide that description, filling in many previous ommissions and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, resulting in the publication of more than 100 peer-reviewed papers based on results obtained with the code. The variety of astrophysical systems that \enzo\ has been used for include galaxies \citep{2003ApJ...587...13T, 2012ApJ...749..140H}, galaxy clusters \citep{Loken02, Skillman13}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01, 2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S}, turbulence \citep{Kritsuk04, 2009JPhCS.180a2020K, Collins11}, the formation of the first stars \citep{ABN02, 2007ApJ...654...66O, 2009Sci...325..601T}, and the formation of stars in our own Galaxy \citep{Collins12a}.
% 'finally' here sounded strange, since this is really the key point
Numerical simulations of astronomical phenomena now play a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01, DecadalSurvey10}.   But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is to document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

### 2013-05-29 numerical-gravity.tex
minor changes to gravity section

@@

First, the dark matter particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  When doing the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids are also added in this way.  In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), dark matter particles from sibling grids and sub grids that lie within the entire gravitating field are used.  This step can involve communication.

@@

For isolated boundary conditions, we use the James method \citep{James77}.  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In either case, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference scheme to obtain accelerations at the cell centers (except if we are using the staggered Zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).  Particle accelerations are obtained using a (linear) CIC interpolation from the grid.

In order to calculate more accurate potentials on subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub grids as before.  To compute the potential on sub grids, however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each sub grid separately.

@@

It is known \red{(REF)} that simply interpolating the potential without feeding it back to higher levels leads to errors in the potential at more refined levels due to the build-up of errors during the interpolation of coarse boundary values.  In addition, neighboring subgrids are not guaranteed to generate the same potential values because of the lack of a coherent potential solve across the whole hierarchy. In an attempt to partially alleviate this problem, we allow for an iterative procedure across sibling grids, in which the potential values on the boundary of grids can be updated with the potential in `active' regions of neighboring subgrids.  To prevent overshoot, we average the potential on the boundary and allow for (typically) 4 iterations.  This procedure can help in many cases, but does not necessarily produce a coherent solution across all grids and so does not completely solve the problem; we are working on a slower but more accurate method that does a multi-grid solve across the whole grid (Reynolds \etal, in preparation).

>>>>

% EJT: small changes only.
% dark matter particles -> massive (dark matter and star) particles

% Changed:  ``Particles from sub-grids are also added in this way'' to ``Particles on sub-grids within the grid's volume are also added to its gravitating field using the same method.'' I expanded because I missed the point and had to clarify with Brian when I read the text. But if it's too long, please just cut it!

% added: ``Each scheme can be selected via the parameter file.'' I felt it sounded like a dice roll otherwise, but again, feel free to overule me!

% removed blank reference [REF]

@@

First, the massive (dark matter and star) particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  During the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids within the grid's volume are also added to its gravitating field using the same method. In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), massive particles from sibling grids and sub grids that lie within the entire gravitating field are used.  This step can involve communication.

@@

For isolated boundary conditions, we use the James method \citep{James77}.  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In both cases, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference scheme to obtain accelerations at the cell centers (except if we are using the staggered Zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).  Particle accelerations are obtained using a (linear) CIC interpolation from the grid.

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub-grids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results but we generally choose to use tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. Each scheme can be selected via the parameter file. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

@@

Simply interpolating the potential without feeding it back to higher levels leads to errors in the potential at more refined levels, due to the build-up of errors during the interpolation of coarse boundary values.  In addition, neighboring subgrids are not guaranteed to generate the same potential values because of the lack of a coherent potential solve across the whole hierarchy. In an attempt to partially alleviate this problem, we allow for an iterative procedure across sibling grids, in which the potential values on the boundary of grids can be updated with the potential in `active' regions of neighboring subgrids.  To prevent overshoot, we average the potential on the boundary and allow for (typically) 4 iterations.  This procedure can help in many cases, but does not necessarily produce a coherent solution across all grids and so does not completely solve the problem; we are working on a slower but more accurate method that does a multigrid solve across the whole grid (Reynolds \etal, in preparation).
